A test of diversifying selection for a trait from within and between species variations

### 2023-06-07 manuscript/main.tex
Updated manuscript

%! BibTeX Compiler = biber
%TC:ignore
\documentclass{article}

\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

@@

\input{notations.tex}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-summary}
    \caption{
        In red, at the phylogenetic scale, the variance of mean trait value ($\VarPhy$) normalised by genetic distance ($d$) is defined as $\RatePhy$.
        In blue, trait variance within species ($\VarGenetic$) normalised by genetic diversity ($\pi$) is defined as $\RatePop$.
        Under neutral evolution $\RatePhy$ should equal $\RatePop$.
        Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
    }
    \label{fig:methods}
\end{figure*}

@@

\subsubsection{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrSpecies$ taxa in the tree, $\Cr$ is the $\NbrSpecies \times \NbrSpecies$ contrast matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
The contrast matrix is symmetric, along the diagonal are the total distances ($d_{\Spi}$) from the root of the tree to each taxon ($\Spi$).
The off-diagonal elements $C_{\Spi,\Spj}$ are the total branch lengths shared by particular pairs of taxa $\Spi$ and $\Spj$, computed from the distance between the pair ($d_{\Spi,\Spj}$) as:
\begin{gather}
    C_{\Spi,\Spj} = d_{\mathcal{A}(\Spi, \Spj)} = \frac{d_{\Spi} + d_{\Spj} - d_{\Spi,\Spj}}{2}
\end{gather}
Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrSpecies \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate\cite{omeara_testing_2006}, given as:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrSpecies \times 1$ column vector of ones.

Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated\cite{omeara_testing_2006} as:
\begin{gather}
    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrSpecies - 1}. \label{eq:estimated-rate-phy}
\end{gather}

For a species $i$ of interest with inter-individual data available, $\RatePopSpi$ can be estimated as the ratio of genetic variance of a trait ($\VarGeneticSpi = \VarPhenotypeSpi \Multiply \HereditySpi$) over genetic diversity of neutrally evolving sequences ($\pi_{\Spi}$).
Averaged across all species, we obtain the mean as:
\begin{gather}
    \EstRatePop = \dfrac{1}{\NbrSpecies}\sum_{i=1}^{\NbrSpecies}\RatePopSpi = \dfrac{1}{\NbrSpecies}\sum_{i=1}^{\NbrSpecies} \frac{  V_{\Trait, i} \Multiply \Heredity_{i}}{ \pi_{i}}. \label{eq:estimated-rate-pop}
\end{gather}

@@

\subsubsection{Bayesian estimate}
We used the software Bayescode

@@

\begin{table*}[t!]
    \centering
    \begin{adjustbox}{width = 1.0\textwidth}
    \begin{tabular}{||l|l||c|c||c|c||}
            \hline
            Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NIx > 1$  & Test $\NIx < 1$ \\ \hline \hline
            Changing $\MutationRate$ & Scaling with $\pi$, $d$, $\VarPhenotype$ and $\VarPhy$ & Adjusted & Adjusted & -- & --  \\ \hline
            Changing $\Ne$ & Scaling with $\pi$ and $\VarPhenotype$ & Adjusted & -- & -- & --  \\ \hline
            Unknown heredity & $\Heredity = 1$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Inbreeding & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Polymorphism are selected & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Sexual dimorphism & $\VarPhenotype$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Trait encoded by few loci & $\VarPhy$ is under-estimated  & -- & Under-estimated & Conservative & False positives  \\ \hline
            Divergence are selected & $d$ is under-estimated & -- & Over-estimated & False positives & Conservative  \\ \hline
            Structured population & unknown & unknown & unknown & unknown & unknown \\ \hline
            Gene flow between species & unknown & unknown & unknown & unknown & unknown \\ \hline
    \end{tabular}
    \end{adjustbox}
    \caption{Assumptions and their consequences on deviation from neutraliity.}
    \label{table:assumptions}
\end{table*}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/constant_pop_size_phy_pop.hist}
    \includegraphics[width=\textwidth, page=1] {artworks/fluctuating_pop_size_phy_pop.hist}
    \caption{
        $\RatePhy / \RatePop$ for $30.000$ genes simulated under different evolutionary regimes.
        $\NIx < 1$ for traits simulated under selection (stabilizing selection in yellow).
        $\NIx = 1$ for traits simulated under a neutral evolution (in blue).
        $\NIx > 1$ for genes simulated under a moving optimum (diversifying selection in red).
        Effective population size ($\Ne$) and mutation rate $\MutationRate$ are either constant (top panel), or fluctuating as a Brownian process along the phylogeny (panel B).
    }
    \label{fig:constant_pop_size_phy_pop}
\end{figure*}

@@

\part*{Supplementary materials}

\section{Theoretical derivation}
\subsection{Definitions}

We denote $\Trait_{i}$ the quantitative trait $\Trait$ for individual $i$.

\begin{align}
    \VarPhenotype = \sum_{i=1}^{\Ne} \Trait_{i}^2 - \left( \sum_{i=1}^{\Ne} \Trait_{i} \right)^2
\end{align}

\subsection{Simulator}
\begin{itemize}
    \setlength\itemsep{-0.2em}
    \item $\NbrLoci$ is the number of loci encoding the trait.
    \item $a_l \sim \mathcal{N}(0,a^2)$ is the effect of a mutation on the trait at locus $l \in \{1, \hdots, \NbrLoci\}$.
    \item $\Ne$ is the effective number of individual in the population.
    \item $g_{i,l} \in \{0, 1, 2\}$ is the genotype at locus $l$ for individual $i \in \{1, \hdots, \Ne\}$.
    \item $G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l}$ is the influence of genotype on the trait for individual $i$.
\end{itemize}
The mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
\begin{equation}
    \bar{G} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} G_i  \text{\quad and \quad}	\VarGenetic = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(G_i - \bar{G} \right)^2
\end{equation}
\begin{itemize}
    \setlength\itemsep{-0.2em}
    \item $\xi_i \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual i.
    \item $\Trait_i = G_i + \xi_i$ is the phenotype for individual $i$.
\end{itemize}
The mean ($\bar{\Trait}$) and variance ($\VarPhenotype$) of the phenotype are:
\begin{equation}
    \bar{\Trait} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} \Trait_i \text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(\Trait_i - \bar{\Trait} \right)^2
\end{equation}
Heritability ($h^2$) is defined as:
\begin{equation}
    h^2 = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}
\end{equation}
\begin{gather}
    G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l} \\
    \Trait_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l} + \xi_i
\end{gather}

@@

\begin{itemize}
    \item $W_i$ is the weight for individual $i$ to be chosen for the next generation.
    \item Neutral model with no fitness function: $W_i = 1$.
    \item Stabilizing selection with fitness $W_i = \e^{(\Trait_i - \theta)^2/ \sigma}$, parameters $\theta=0$ and $\sigma=0.02$.
    \item Moving optimum with fitness $W_i = \e^{(\Trait_i - \theta_t)^2/ \sigma}$, parameters $\theta_t \sim \mathcal{B}_t$ and $\sigma=0.02$.
\end{itemize}
\begin{itemize}
    \item Trait encoded by $5,000$ loci, $1000$ replicates.
    \item $a=1$, $h^2=0.2$, $\Ne=50$, $\mu=10^{-5}$, $t=10,000$.
\end{itemize}

\begin{table*}[t!]
    \centering
    \begin{adjustbox}{width = 0.5\textwidth}
        \begin{tabular}{|c||c|c|}
            \hline
            & Coding sequences & Quantitative traits \\ \hline \hline
            Adaptation / Diversifying selection & $\dnds > \pnps$ & $ \RatePhy > \RatePop  $ \\ \hline
            Neutral regime of evolution & $\dnds = \pnps$ & $ \RatePhy = \RatePop  $ \\ \hline
            Purifying / Stabilizing selection & $\dnds < \pnps$ & $ \RatePhy < \RatePop  $ \\ \hline
        \end{tabular}
    \end{adjustbox}
    \caption{
        The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
        Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
    }
    \label{table:dnds}
\end{table*}

\end{document}
%TC:endignore

>>>>

%! BibTeX Compiler = biber
%TC:ignore
\documentclass{article}
\usepackage{caption}
\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

@@

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heredity}{h^2}
\newcommand{\HereditySpi}{\Heredity_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\widehat{\theta}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{G}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{G}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\MatrixGenetic}{\BiDimArray{G}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RatePhy}{\sigma^2_{\mathrm{B}}}
\newcommand{\RatePop}{\sigma^2_{\mathrm{W}}}
\newcommand{\RatePopSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRatePop}{\UniDimArray{\RatePop}}
\newcommand{\EstRatePhy}{\widehat{\RatePhy}}
\newcommand{\EstRatePop}{\widehat{\RatePop}}
\newcommand{\NI}{\rho}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

% Tree
\newcommand{\branch}{\text{b}}
\newcommand{\branchexp}{^{(\branch)}}
\newcommand{\Nbranch}{2 \NbrTaxa - 2}
\newcommand{\WishartPostDf}{2 \NbrTaxa + 1}
\newcommand{\Setbranch}{\branch \in \{1, \hdots, \Nbranch \}}
\newcommand{\nodeUp}{\branch^{\uparrow}}
\newcommand{\nodeDown}{\branch^{\downarrow}}
\newcommand{\Ntrait}{K}
\newcommand{\contrast}{\UniDimArray{C}}
\newcommand{\Covariancematrix}{\Sigma}
\newcommand{\CovarianceMatrix}{\BiDimArray{\Covariancematrix}}
\newcommand{\Precisionmatrix}{\Omega}
\newcommand{\PrecisionMatrix}{\BiDimArray{\Precisionmatrix}}
\newcommand{\Identitymatrix}{\BiDimArray{I}}
\newcommand{\brownian}{\mathcal{B}}
\newcommand{\Brownian}{\UniDimArray{\brownian}}
\newcommand{\Scattermatrix}{\BiDimArray{A}}
\newcommand{\Multivariate}{\UniDimArray{Z}}
\newcommand{\indice}{a}
\newcommand{\indiceexp}{^{(\indice)}}
\newcommand{\vecZero}{\UniDimArray{0}}
\newcommand{\vecOne}{\UniDimArray{1}}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
    \caption{
        Between species, the variance of mean trait value ($\VarPhy$) normalised by genetic distance ($d$) is defined as $\RatePhy$.
        Within species, genetic variance ($\VarGenetic$) normalised by genetic diversity ($\pi$) is defined as $\RatePop$.
        Under neutral evolution $\RatePhy$ should equal $\RatePop$.
        Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
    }
    \label{fig:methods}
\end{figure*}

@@

\subsubsection{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
Along the diagonal $\Distance_{\Spi,\Spi}$ are the total distances from the root of the tree to each taxon ($\Spi$).
The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be root, in which case the distance is 0).
Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate\cite{omeara_testing_2006}, given as:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.

Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated\cite{omeara_testing_2006} as:
\begin{gather}
    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}. \label{eq:estimated-rate-phy}
\end{gather}

For a given species $\Spi$ with inter-individual data available, the ratio of genetic variance of a trait ($\VarGeneticSpi$) over genetic diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RatePop$.
Averaged across all species, we obtain the estimate $\EstRatePop$ as:
\begin{gather}
    \EstRatePop = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heredity_{i}}{ \pi_{i}}. \label{eq:estimated-rate-pop}
\end{gather}

@@

\subsubsection{Bayesian estimate}

Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them.
Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
We defined the {prior} on the covariance matrix as an inverse Wishart distribution, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ is also an invert Wishart distribution.

@@

\begin{table*}[t!]
    \centering
    \begin{adjustbox}{width = 1.0\textwidth}
    \begin{tabular}{||l|l||c|c||c|c||}
            \hline
            Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NI > 1$  & Test $\NI < 1$ \\ \hline \hline
            Changing $\MutationRate$ & Scaling with $\pi$, $d$, $\VarPhenotype$ and $\VarPhy$ & Adjusted & Adjusted & -- & --  \\ \hline
            Changing $\Ne$ & Scaling with $\pi$ and $\VarPhenotype$ & Adjusted & -- & -- & --  \\ \hline
            Unknown heredity & $\Heredity = 1$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Inbreeding & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Polymorphism are selected & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Sexual dimorphism & $\VarPhenotype$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Trait encoded by few loci & $\VarPhy$ is under-estimated  & -- & Under-estimated & Conservative & False positives  \\ \hline
            Divergence are selected & $d$ is under-estimated & -- & Over-estimated & False positives & Conservative  \\ \hline
            Structured population & unknown & unknown & unknown & unknown & unknown \\ \hline
            Gene flow between species & unknown & unknown & unknown & unknown & unknown \\ \hline
    \end{tabular}
    \end{adjustbox}
    \caption{Assumptions and their consequences on deviation from neutraliity.}
    \label{table:assumptions}
\end{table*}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.rho}
    \includegraphics[width=\textwidth, page=1] {artworks/cst_L5000.pvalues}
    \caption{
        $\RatePhy / \RatePop$ for $1,000$ replicates under different evolutionary regimes.
        Traits are encoded by $5,000$ loci with $a=1$, $h^2=0.2$, $\Ne=50$, $\mu=10^{-5}$, $t=10,000$.
        $\NI < 1$ for traits simulated under selection (stabilizing selection in yellow).
        $\NI = 1$ for traits simulated under a neutral evolution (in blue).
        $\NI > 1$ for genes simulated under a moving optimum (diversifying selection in red).
        Effective population size ($\Ne$) and mutation rate $\MutationRate$ are either constant (top panel), or fluctuating as a Brownian process along the phylogeny (panel B).
    }
    \label{fig:constant_pop_size_phy_pop}
\end{figure*}

@@

@@

\part*{Supplementary materials}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\setcounter{table}{0}

\section{Simulator}

\subsection{Genotype-phenotype map}

\begin{itemize}
    \setlength\itemsep{-0.2em}
    \item $\NbrLoci$ is the number of loci encoding the trait.
    \item $a_l \sim \mathcal{N}(0,a^2)$ is the effect of a mutation on the trait at locus $l \in \{1, \hdots, \NbrLoci\}$.
    \item $\Ne$ is the effective number of individual in the population.
    \item $g_{i,l} \in \{0, 1, 2\}$ is the genotypic value at locus $l$ for individual $i \in \{1, \hdots, \Ne\}$.
    \item $G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l}$ is the genotypic value for individual $i$.
\end{itemize}

\begin{center}
    \captionof{figure}{Summary of the simulator.}
    \includegraphics[width=0.6\textwidth, page=1] {artworks/fig-simulator-model}
    \label{fig:simulator-summary}
\end{center}

\subsection{Heritability, genotypic and phenotypic variance}
At the population level, the mean ($\bar{G}$) and variance ($\VarGenetic$) of the genotype are:
\begin{equation}
    \bar{G} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} G_i  \text{\quad and \quad}	\VarGenetic = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(G_i - \bar{G} \right)^2
\end{equation}
\begin{itemize}
    \setlength\itemsep{-0.2em}
    \item $\xi_i \sim \mathcal{N}(0, \VarEnv)$ is the effect of environment on the trait for individual i.
    \item $\Trait_i = G_i + \xi_i$ is the phenotype for individual $i$.
\end{itemize}
The mean ($\bar{\Trait}$) and variance ($\VarPhenotype$) of the phenotype are:
\begin{equation}
    \bar{\Trait} = \frac{1}{\Ne}\sum_{i=1}^{\Ne} \Trait_i \text{\quad and \quad} \VarPhenotype = \frac{1}{\Ne}\sum_{i=1}^{\Ne}\left(\Trait_i - \bar{\Trait} \right)^2
\end{equation}
Heritability ($h^2$) is defined as:
\begin{equation}
    h^2 = \frac{\VarGenetic}{\VarPhenotype} = \frac{\VarGenetic}{\VarGenetic + \VarEnv}
\end{equation}
\begin{gather}
    G_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l} \\
    \Trait_i = \sum_{l=1}^{\NbrLoci} a_l \times g_{i,l} + \xi_i
\end{gather}

\subsection{Phenotype-fitness map}

\begin{itemize}
    \item $W_i$ is the weight for individual $i$ to be chosen for the next generation.
    \item Neutral model with no fitness function: $W_i = 1$.
    \item Stabilizing selection with fitness $W_i = \e^{(\Trait_i - \theta)^2/ \sigma}$, parameters $\theta=0$ and $\sigma=0.02$.
    \item Moving optimum with fitness $W_i = \e^{(\Trait_i - \theta_t)^2/ \sigma}$, parameters $\theta_t \sim \mathcal{B}_t$ and $\sigma=0.02$.
\end{itemize}

\section{Analogy with molecular evolution}

\begin{center}
    \captionof{table}{Analogy with molecular evolution}
    \scriptsize
    \begin{adjustbox}{width = 0.5\textwidth}
        \begin{tabular}{|c||c|c|}
            \hline
            & Coding sequences & Quantitative traits \\ \hline \hline
            Adaptation / Diversifying selection & $\dnds > \pnps$ & $ \RatePhy > \RatePop  $ \\ \hline
            Neutral regime of evolution & $\dnds = \pnps$ & $ \RatePhy = \RatePop  $ \\ \hline
            Purifying / Stabilizing selection & $\dnds < \pnps$ & $ \RatePhy < \RatePop  $ \\ \hline
        \end{tabular}
    \end{adjustbox}
\end{center}
The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.

\section{Bayesian estimate}

\subsection{Multivariate Brownian process}
Here we generalize to $\Ntrait$  traits evolving along the phylogeny and are correlated between them.
Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology.
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
For a branch $\branch$ with length $d\branchexp$, the Brownian process start at node $\nodeUp$ with value $\Brownian(\nodeUp)$, at ends at node $\nodeDown$ with value $\Brownian(\nodeDown)$.
The independent contrast $\contrast\branchexp$ defined as change in trait along the branch normalized by $\sqrt {d}$ is a multivariate Gaussian:
\begin{equation}
    \label{eq:DistribBrownian}
    \contrast\branchexp = \frac{\Brownian^{(\nodeDown)} - \Brownian^{(\nodeUp)} }{\sqrt {d}} \sim \mathcal{N}\left(\vecZero, \CovarianceMatrix \right).
\end{equation}
We defined the {prior} on the covariance matrix as an inverse Wishart distribution, with $\Ntrait + 1$ degrees of freedom:
\begin{equation}
    \label{eq:Distribcovariance}
    \CovarianceMatrix \sim \text{Wishart}^{-1} (\Identitymatrix, \Ntrait + 1).
\end{equation}

\subsection{Sampling the covariance matrix}
From the independent contrast at each branch of the tree ($\contrast\branchexp$), we can define the $\Ntrait \times \Ntrait$ scatter matrix, $\Scattermatrix$ as:
\begin{equation}
    \Scattermatrix = \sum\limits_{b=1}^{\Nbranch} \contrast\branchexp \MultiplyMatrix \left[\contrast\branchexp\right]\tr
\end{equation}
By Bayes theorem, the {posterior} on $\CovarianceMatrix$, conditional on a particular realization of $\brownian$ (and thus of $\contrast$) is an invert Wishart distribution, of parameter $\Identitymatrix + \Scattermatrix$ and with $\WishartPostDf$ degrees of freedom.
\begin{equation}
    \CovarianceMatrix \sim \text{Wishart}^{-1}\left( \Identitymatrix + \Scattermatrix, \WishartPostDf\right)
\end{equation}
This invert Wishart distribution can be obtained by sampling $\WishartPostDf$ independent and identically distributed multivariate normal random variables $\Multivariate\indiceexp$ defined by
\begin{equation}
    \Multivariate\indiceexp \sim \mathcal{N} \left( \vecZero, \left[ \Identitymatrix + \Scattermatrix\right]^{-1} \right).
\end{equation}
And from these multivariate samples, $\CovarianceMatrix$ is Gibbs sampled as:
\begin{equation}
    \CovarianceMatrix = \left( \sum\limits_{k=1}^{\WishartPostDf} \Multivariate\indiceexp \MultiplyMatrix  \left[\Multivariate\indiceexp \right] \tr \right)\inv
\end{equation}

\subsection{Implementation}
\label{sec:site-specific-mutation-selection-codon-models}
In \textit{BayesCode} (\url{https://github.com/ThibaultLatrille/bayescode}), we ran the Brownian multivariate model \textit{nodetraits} for 2000 points of MCMC with the options:
\begin{scriptsize}
    \begin{verbatim}
 nodetraits --tree my_tree.newick --traits mean_traits.tsv -u 2000 run_name_chain1
    \end{verbatim}
\end{scriptsize}
The covariance matrix ($\CovarianceMatrix$) is also obtained by running \textit{readnodetraits}, reading 1000 points of MCMC (first 1000 are considered as burn-in) with the options:
\begin{scriptsize}
    \begin{verbatim}
 readnodetraits --every 1 --until 2000 --burnin 1000 --cov run_name_chain1
    \end{verbatim}
\end{scriptsize}
\end{document}
%TC:endignore
